---
title: "Final Project Quarto Analysis"
format: html
---

---
title: "Final Project Quarto Analysis"
format: html
---
 
Load in necessary packages 
```{r, message=FALSE}
library(tidyverse)
library(janitor)
library(here)
source(here("R", "moving_average.R")) # call a function you will be using 
source(here("R", "chemical_labeller.R")) # call a function you will be using 
```

# Background 
This analysis plots the chemical concentration of samples taken from various stream water sites in the Luquillo Mountains, Puerto Rico. 

It calculates the 9 week moving average for every chemical concentration and plots it across time. From this plot, the effects of Hurricane Hugo on stream chemical concentrations can be observed. 

# Data
Chemical concentration data comes from the following Bisley stream water samples sites, Quebrada Cuenca site 1, 2, and 3 (q1, q2, q3) and Puente Roto Mameyes (PMR). The column names have been cleaned, the site tables have been joined, and the dates filtered for the desired date range. 
```{r, message=FALSE, warning=FALSE}
# Source an intermediate file containing the cleaned and wrangled data frame.
source(here("1_df_clean_wrang.R"))
```


# Methods
The following code applies a moving average function to the data frame and inserts this  calculation into a new column called "chemical name"_moving_avg. 
```{r}
# Create potassium moving average column 
joined_sites_avg <- joined_sites %>% 
  group_by(sample_id) %>% 
  mutate(k_moving_avg = sapply(
    sample_date,
    moving_average,
    dates = sample_date,
    conc = k,
    win_size_wks = 9
  )) %>% 
  ungroup()

# Create dissolved nitrogen moving average column 
joined_sites_avg <- joined_sites_avg %>% 
  group_by(sample_id) %>% 
  mutate(no3_n_moving_avg = sapply(
    sample_date,
    moving_average,
    dates = sample_date,
    conc = no3_n,
    win_size_wks = 9
  )) %>% 
  ungroup()

# Create magnesium moving average column 
joined_sites_avg <- joined_sites_avg %>% 
  group_by(sample_id) %>% 
  mutate(mg_moving_avg = sapply(
    sample_date,
    moving_average,
    dates = sample_date,
    conc = mg,
    win_size_wks = 9
  )) %>% 
  ungroup()

# Create calcium moving average column 
joined_sites_avg <- joined_sites_avg %>% 
  group_by(sample_id) %>% 
  mutate(ca_moving_avg = sapply(
    sample_date,
    moving_average,
    dates = sample_date,
    conc = ca,
    win_size_wks = 9
  )) %>% 
  ungroup()

# Create ammonium nitrogen moving average column 
joined_sites_avg <- joined_sites_avg %>% 
  group_by(sample_id) %>% 
  mutate(nh4_n_moving_avg = sapply(
    sample_date,
    moving_average,
    dates = sample_date,
    conc = nh4_n,
    win_size_wks = 9
  )) %>% 
  ungroup()

```


This code uses pivot longer to place all chemical types into their own column called "chemical" and their concentration average in a separate this column. 

This step will allow the use of facet_grid which relies on a single column to create individual panels. 
```{r}
# Pivot longer 
long_sites_avg <- joined_sites_avg %>% 
  pivot_longer(k_moving_avg:nh4_n_moving_avg,
               names_to = "chemical", values_to = "concentration_average")
```


This code produces a plot with separate panels for each chemical. Each panel contains the chemical concentration over time with distinct lines for each site. 
```{r, warning=FALSE}
# Create a list which contains the replacement names for the previous facet grid x axis strip labels. 
chemical_names <- list(
  'k_moving_avg'="K mg l^-1",
  'no3_n_moving_avg'="NO3-N ug l^-1",
  'mg_moving_avg'="Mg mg l^-1",
  'ca_moving_avg'="Ca mg l^-1", 
  'nh4_n_moving_avg' = "NH4-N ug l-1"
)

# Plot
chem_site_plot <- long_sites_avg %>% 
  ggplot(aes(x = sample_date, y = concentration_average, color = sample_id)) +
  geom_line(size = 1) +
  scale_color_discrete(labels = c("PMR", "BQ1", "BQ2", "BQ3")) + # change labels in legend
  facet_grid(rows = vars(factor(chemical,
                     levels=c("k_moving_avg", "no3_n_moving_avg", "mg_moving_avg", "ca_moving_avg", "nh4_n_moving_avg"))), scales = "free", switch = "y", labeller=chemical_labeller) + # apply labeller function to replace the plots x axis title.  
  theme_classic() + 
  theme(legend.position = c(1.07, 0.9), # adjust plot sections
        plot.margin = unit(c(1,2.4,0.5,0.4), "cm"),
        axis.title.y = element_blank(), 
        panel.border = element_rect(color = "black", 
                                    fill = NA, size = 1),
        plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(size = 13), 
        legend.title=element_blank(),
        plot.caption = element_text(size = 10, hjust = 0)
        ) +
  xlab("Years") + 
  labs(title = "Years", caption = "Figure 3. Concentrations in Bisley, Puerto Rico streams before and after Hurricane Hugo, 9-wk moving 
  averages.(a) potassium, (b) nitrate-N, (c) magnesium, (d) calcium and (e) ammonium-N. The vertical 
  lines mark the time of hurricane disturbance.") + 
  geom_vline(xintercept = as.Date("1989-09-19"), linetype = 3) # insert vertical line indicating the occurance of hurricane hugo
  
```

# Results
Chemical concentrations of different sites across time. 

The occurrence of Hurricane Hugo is visulaized with a horizontal line, it demonstrates how chemical concentrations changed after it.
```{r, fig.width = 8, fig.height = 10, warning= FALSE, echo = FALSE}
chem_site_plot # Output the final plot 
```



