---
title: "Final Project Quarto Analysis"
format: html
---

---
title: "Final Project Quarto Analysis"
format: html
---
 
Load in necessary packages 
```{r, message=FALSE}
library(tidyverse)
library(janitor)
library(here)
source(here("R", "moving_average.R"))
source(here("R", "chemical_labeller.R"))
```

# Background 
This analysis plots the chemical concentration of samples taken from various stream water sites in the Luquillo Mountains, Puerto Rico. 

It calculates the 9 week moving average for every chemical concentration and plots it across time. From this plot, the effects of Hurricane Hugo on chemical concentrations can be observed. 

# Data
Chemical concentration data comes from the following Bisley stream water samples sites, Quebrada Cuenca site 1, 2, and 3 (q1, q2, q3) and Puente Roto Mameyes (PMR). The column names have been cleaned, the site tables have been joined, and the dates filtered for the desired date range. 

```{r, message=FALSE, warning=FALSE}
source(here("1_df_clean_wrang.R"))
```

